---
title: "Biobest Dataset Analysis"
author: "Author: Catarina Gouveia"
output: html_document
theme: united
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(readxl)
library(knitr)
```


# Tasks: {.tabset .tabset-pills}
<br/>
1. Descriptive analytics
    - Describe the dataset in comprehensive figures/tables. 
    What can we learn from this dataset? What do you see? 
<br/>
<br/>
2. Predictive analytics
    - Use your statistical skills to build a first predictive model (e.g. regression models, differential equations, AI,…)
    - Question: can we predict the population dynamics one or two weeks in advance??
<br/>
<br/>
3. Discussion: What would you do with this dataset if you had more time? Which techniques could improve the model? Are we missing parameters?
<br/>

## Data Exploration 
```{r}
data <- read_excel("~/Documents/DataScienceJobsChallenges/DataScienceProjects/DataScienceProjects/BiobestData/Test dataset.xlsx")
nrow(data)
ncol(data)
colnames(data)
str(data)
summary(data)
```

### Check data

```{r}
head(data)
```
```{r}
data$Plant <- as.numeric(data$Plant)
data$Prey <- as.numeric(data$Prey)
data$Predator <- as.numeric(data$Predator)
data$Datum <- as.Date(data$Datum)
# Transform week nr to date to see if it matches:
data$WeekDate <- as.Date(paste(2016, data$Week, 2, sep="-"), "%Y-%U-%u")
```

(Note: there is a slight desilign but the week was probably generated from the 
Datum)<br/>

### Check nr of unique preys, plants and predators<br/>
```{r}
nrow(unique(data["Prey"]))
nrow(unique(data["Plant"]))
nrow(unique(data["Predator"]))
```
```{r}
unique(data$Row)
```

```{r,message=FALSE, warning=FALSE}
library(dplyr)
# list rows with missing values
data %>% 
  filter(if_any(everything(), is.na))
```
330 rows with missing Prey and Predator data<br/><br/>

Continuing exploration:<br/><br/>

Count number of rows with missing data per plant:<br/>
```{r}
data %>% 
  filter(if_any(everything(), is.na)) %>% 
  count(Plant)
```
Since plants 31,32,33,34,35 are missing a lot of data, we can remove them \
from the final analysis (they might be considered as outliers).
But before that let's look into more detail to plants 34 and 35 to confirm \
that we can remove them.<br/>
```{r}
data %>% filter(Plant ==34)

data %>% filter(Plant ==35)
```

How many data points were collected each week?<br/>
```{r, echo = FALSE}
table(data$Week)
```
- 315 each week<br/><br/>

Let's look into which points are collected each week. Are they the same?<br/>
```{r}
data %>% 
  filter(Week==18)

# Because there are only 35 plants check what are the data points for week 18, plant 1
data %>% 
  filter(Week==18 & Plant == 1)
# And now plant 2
data %>% 
  filter(Week==18 & Plant == 2)

table(data$Plant)

table(data$Plant, data$Row)
```

### - Prey: Number of insects/bugs in the plant<br/>
### - Predator: Number of predators for the preys in the plant<br/>
### - Row: Location where plant is physically<br/>
### - Week: Data collection date point<br/>
</br>
Notes: <br/>
- There are 9 records per plant, per week (9 weeks, starting at 2016-05-03, 18th week of the year)<br/>
- There are 9 rows, with 35 plants each<br/>


### To better understand the data:<br/>

#### 1. Check the progression of preys and predators along the weeks per row (all plants in one row, together)
```{r echo=FALSE, warning=FALSE, message=F, fig.dim=c(12, 40), cache=T}
library(gridExtra)
library(RColorBrewer)
library(grid)
library(ggpubr)
library(tidyr)



d <- data %>%
  # Create counts column that holds nr of preys and predators found
    gather(Prey, Predator, key = "Prey/Predator", value = "Counts")


# Plot
for (var in 1:length(unique(data$Plant))){
  assign(paste0("gg_", var), ggplot(d %>% 
             filter(Plant == var)
             ) + 
  geom_point(aes(x=Week, y = Row, size = Counts, color=`Prey/Predator`, alpha=0.4)) + 
    scale_color_manual(name="Prey/Predator",
                       breaks=c("Prey", "Predator"),
                       values = c("Prey" = "#66C2A5","Predator"= "#FC8D62")) +
    scale_alpha(guide = 'none') +
  # geom_point(aes(x=Week, y = Row, size = Prey), color="#66C2A5", alpha=0.4, show.legend=F) +
  # geom_point(aes(x=Week, y = Row, size = Predator), color="#FC8D62",alpha=0.4, show.legend=F) +
  # ggtitle(paste("Preys in plant", var)) +
  labs(
       y="Rows", 
       x=paste0("Plant #", var)) + 
  theme(plot.title = element_text(hjust = 0.5),
        plot.margin = margin(1,1,1.5,1.2, "cm"))
  +theme_classic())
  
}


p <- grid.arrange(
  gg_1+ theme(legend.position="none"),
  gg_2+ theme(legend.position="none"),
  gg_3+ theme(legend.position="none"),
  gg_4+ theme(legend.position="none"),
  gg_5+ theme(legend.position="none"),
  gg_6+ theme(legend.position="none"),
  gg_7+ theme(legend.position="none"),
  gg_8+ theme(legend.position="none"),
  gg_9+ theme(legend.position="none"),
  gg_10+ theme(legend.position="none"),
  gg_11+ theme(legend.position="none"),
  gg_12+ theme(legend.position="none"),
  gg_13+ theme(legend.position="none"),
  gg_14+ theme(legend.position="none"),
  gg_15+ theme(legend.position="none"),
  gg_16+ theme(legend.position="none"),
  gg_17+ theme(legend.position="none"),
  gg_18+ theme(legend.position="none"),
  gg_19+ theme(legend.position="none"),
  gg_20+ theme(legend.position="none"),
  gg_21+ theme(legend.position="none"),
  gg_22+ theme(legend.position="none"),
  gg_23+ theme(legend.position="none"),
  gg_24+ theme(legend.position="none"),
  gg_25+ theme(legend.position="none"),
  gg_26+ theme(legend.position="none"),
  gg_27+ theme(legend.position="none"),
  gg_28+ theme(legend.position="none"),
  gg_29+ theme(legend.position="none"),
  gg_30+ theme(legend.position="none"),
  gg_31+ theme(legend.position="none"),
  gg_32+ theme(legend.position="none"),
  gg_33+ theme(legend.position="none"),
  gg_34+ theme(legend.position="none"),
  gg_35+ theme(legend.position="none"),
  get_legend(gg_35 + 
    scale_color_manual(name="Prey/Predator",
                       breaks=c("Prey", "Predator"),
                       values = c("Prey" = "#66C2A5","Predator"= "#FC8D62")) +
    theme(legend.position="right", legend.box = "horizontal",
          legend.title=element_text(size=14),
          legend.text=element_text(size=10))),
  nrow = 12,
  ncol=3,
  top = textGrob("Preys and predators in rows per plant", hjust=0.5,
                 gp=gpar(fontsize=20,font=3)),
  bottom = textGrob(
    "Data Source: Biobest",
    gp = gpar(fontface = 3, fontsize = 9),
    hjust = 1,
    x = 1
  )
) 

# plot(p )

```


```{r, fig.dim=c(20, 40), eval=F, echo=F, cache=T}

var=1
for (row in unique(data$Row)){
  assign(paste0("gg_", var), ggplot(d) + 
  geom_point(aes(x=Week, y = Row, 
                  # group="Prey/Predator",
                 size=Counts, 
                 color="Prey/Predator", 
             alpha=0.4)) +
  scale_color_manual(values = c("Prey" = "#66C2A5","Predator"= "#FC8D62")) +
  labs(
       y="Rows", 
       x="Weeks") + 
  theme(plot.title = element_text(hjust = 0.5),
        plot.margin = margin(1,1,1.5,1.2, "cm")) + 
    theme_classic())
  var<-var+1
}

p <- grid.arrange(
  gg_1 + theme(legend.position="none"),
  gg_2 + theme(legend.position="none"),
  gg_3 + theme(legend.position="none"),
  gg_4 + theme(legend.position="none"),
  gg_5 + theme(legend.position="none"),
  gg_6 + theme(legend.position="none"),
  gg_7 + theme(legend.position="none"),
  gg_8 + theme(legend.position="none"),
  gg_9 + theme(legend.position="right", legend.box = "horizontal"),
  nrow = 3,
  ncol=3,
  top = textGrob("Progression of preys and predators along the weeks",
                 gp=gpar(fontsize=20,font=3)),
  bottom = textGrob(
    "Source: Biobest",
    gp = gpar(fontface = 3, fontsize = 9),
    hjust = -0.5,
    x = 1
  )
)

# ggsave("PlantsPlot_Week.pdf", p, "pdf", width = 40, height = 60, units = "cm")

# plot(p)

```

#### 2. Visualize progression of preys and predators in the dataset, with time points, for all plants
```{r echo=FALSE, warning=FALSE, message=F, cache=T}

# Scatterplot
g <- ggplot(d) +
  geom_point(aes(x=Week, y = `Counts`, color=`Prey/Predator`), 
             size=1, alpha=0.4) +
  labs(
       y="Prey/Predator Counts", 
       x=paste0("Weeks")) + 
  ggtitle("Number of preys and predators")+
  theme(plot.title = element_text(hjust = 0.5),
        plot.margin = margin(1,1,1.5,1.2, "cm")) +
  theme_classic()

# ggsave("PlantsPlot2.pdf", g, "pdf", width = 50, height = 100, units = "cm")

plot(g)
```

```{r, fig.dim=c(40, 80), eval=F, echo=F}

knitr::include_graphics("PlantsPlot2.pdf")

```
<br/>
<br/>
### Another way of depicting the above trends 
Note: I opted for merging plants by number so the data would be visible in a simpler way, however, be aware of this for further analysis
```{r, warning=F, message=F, cache=T}
# Count number of total preys and predators per week
t <- data %>% 
   group_by(Week, Plant) %>% 
   mutate(TotalPreyPerWeek = sum(Prey),
          TotalPredatorPerWeek = sum(Predator)) %>%
   ungroup 

leg_p <- ggplot(t , aes(Week)) +
  geom_line(aes(y=TotalPreyPerWeek, 
                group=Plant, 
            color="dark green"),
            size=1) +
  geom_line( aes(y=TotalPredatorPerWeek, group=Plant, 
            color="dark red"), 
            size=1) +
  scale_color_manual(
    breaks=c("TotalPreyPerWeek", "TotalPredatorPerWeek"),
    values = c("TotalPreyPerWeek" = "dark green",
               "TotalPredatorPerWeek"= "dark red")) +
 theme(legend.title=element_text(size=20),
       legend.text=element_text(size=14)) +
  theme_classic()

p1 <- ggplot(t , aes(Week)) +
  geom_line(aes(y=TotalPreyPerWeek, 
                group=Plant
            ), color="dark green",
            size=1) +
  geom_line( aes(y=TotalPredatorPerWeek, group=Plant
            ), color="dark red",
            size=1) +
  scale_color_manual(
    breaks=c("TotalPreyPerWeek", "TotalPredatorPerWeek"),
    values = c("TotalPreyPerWeek" = "dark green",
               "TotalPredatorPerWeek"= "dark red")) +
 theme(legend.title=element_text(size=20),
       legend.text=element_text(size=14)) +
  theme_classic() + ggtitle("Prey and Predator Distribution/Trend")

grid.arrange(p1,
             get_legend(leg_p), 
             nrow=1, 
             ncol=2, 
             widths=c(3,1))

```
<br/><br/>Another possible option:
3. Visualize progression of preys and predators in the dataset, for all plants, each week separately (less important)


### Some conclusions:

#### - By looking at the data, it looks like the number of preys grows until week 22 
(starting on week 18) and then it starts to decrease, with the number of 
predators reaching its highest on weeks 23 and 24 (then decreasing as well)


## Predictive Analytics 
- Depict what is going to happen on week 27 and 28

- Try to predict what is going to happen on week 27 and 28

### Lotka-Volterra Model
```{r}
library(deSolve)
```

```{r}
## -----------------------------------------------------------------------------
## Define R-function
## -----------------------------------------------------------------------------   

LV <- function(t, y, parms) {
    with(as.list(c(y, parms)), {
  
        # dP <- rG * P * (1 - P/K) - rI * P * C # Prey
        # dC <- rI * P * C * AE - rM * C # Predator
        dP <- alpha * P - beta * C * P # Prey
        dC <- gamma * P * C - delta * C # Predator
            
        return(list(c(dP, dC), sum = C+P))
    })
}

# Source1: https://riptutorial.com/r/example/24636/lotka-volterra-or--prey-vs--predator
# Source2: https://www.sciencedirect.com/topics/earth-and-planetary-sciences/lotka-volterra-model
```

```{r}

s <- data %>% 
   group_by(Week, Plant) %>% 
   mutate(TotalPreyPerWeek = mean(Prey),
          TotalPredatorPerWeek = mean(Predator)) %>%
  slice(1) %>%
   ungroup 

## -----------------------------------------------------------------------------
## Define parameters and variables
## -----------------------------------------------------------------------------

# As a toy example let's use the Total Prey or Predator Per Week calculated above, as a starting point:
startPrey <-
  s %>% 
  filter(Plant == 1) %>% 
  select(TotalPreyPerWeek) %>% 
  head(1) %>% 
  as.numeric()

startPredator <-
  s %>% filter(Plant == 1) %>% 
  select(TotalPredatorPerWeek) %>% 
  head(1) %>% 
  as.numeric()

startDate <-
  s %>% filter(Plant == 1) %>% 
  select(Week) %>% head(1) %>% 
  as.numeric()

endDate <-
  s %>% filter(Plant == 1) %>% 
  select(Week) %>% tail(1) %>% 
  as.numeric() + 4 # + 4 weeks

State <- c(P = startPrey, C = startPredator+3)
Time <- seq(startDate, endDate, by = 1)


# parms <- c(rI = 0.2, rG = 1.0, rM = 0.2, AE = 0.5, K = 10)

# The prey grows at a linear rate (alpha) and gets eaten by the predator at the rate of (beta). The predator gains a certain amount vitality by eating the prey at a rate (delta), while dying off at another rate (gamma).

parms <-
  c(
    alpha = 1.5,
    beta = .6,
    gamma = 1.2,
    delta = 0.8
  ) # These are rough estimations

# The above parameters can be improved


out <-
  as.data.frame(ode(
    func = LV,
    y = State,
    parms = parms,
    times = Time
  ))

matplot(out[, -1],
        type = "l",
        xlab = "Time",
        ylab = "Population")
legend(
  "topright",
  c("Preys", "Predators"),
  lty = c(1, 2),
  col = c(1, 2),
  box.lwd = 0)
  

```


Note: To be properly done:

1. Run a logistic(?) model
2. Get the coeficients of the model (with coef())
3. Use the right parameters and coefficients in the right function
4. Predict what is going to happen further in time
